---
title: "College Football Regression"
author: "Blake Behrens"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
scores = read.csv('scores.csv')
postszn = read.csv("bowls25.csv")
```

# Linear Regression in College Football Prediction

The use of linear regression models in sports prediction is well documented and proved as a consistent and overall reliable model. The goal of this project is to analyze which predictors do best in forecasting the margin of scores between home team and away team. For this research, the regular season data will be built to train the model which will then be evaluated on playoff and bowl games (2025 season only).

## Basic Model

The first, most basic model to be fit is a multiple linear regression where observed score margin is explained by a difference in 2 team strengths. 

$$M_{ij} = S_i - S_j + \epsilon_{ij} $$
Where:

- $M_{ij}$ is the observed scoring margin for team *i* playing at home against team *j*

- $S_i,S_j$ are strengths of team *i* and *j* respectively

- $\epsilon_{ij}$ is the unobserved, mean-zero random error for the game between *i* and *j*

### Rankings: 

Below is this model's top 25 rankings, with the strength estimate representing the expected margin of victory against a true average team.


```{r Basic Model set-up, echo=FALSE}
basic.df = scores[scores$FBS == 1,c("home","away","margin")]
N = nrow(basic.df)
teams = unique(basic.df$home)
nteams = length(teams)
home = basic.df$home
away = basic.df$away

X = matrix(0, nrow = N, ncol = nteams)
i = 1
for(team in teams){
X[,i] = (home == team) -
  (away == team)
i = i + 1
}

y = basic.df$margin
basic.model = lm(y ~ X + 0)
b = coef(basic.model)
b = b - mean(b)
basic.ranks = data.frame("Team" = teams, "Strength" = b, row.names = 1:nteams)
head(arrange(basic.ranks,-Strength),25)
```


### Playoff Accuracy

```{r echo = F}
basic.evaluation = postszn %>% 
  mutate(margin = h.score - a.score) %>% 
  left_join(basic.ranks, by = join_by(home == Team)) %>% 
  dplyr::select(home,away,margin,homestrength = Strength) %>% 
  left_join(basic.ranks, by = join_by(away == Team)) %>% 
  dplyr::select(home,away,margin,homestrength, awaystrength = Strength) %>% 
  mutate(pred = homestrength - awaystrength) %>% 
  filter(!is.na(homestrength))

accuracy = mean(sign(basic.evaluation$margin) == sign(basic.evaluation$pred))
MSE = mean((basic.evaluation$margin - basic.evaluation$pred)^2)
r2 = cor(basic.evaluation$margin, basic.evaluation$pred)^2
cat("Basic Model: ", "\n",
    "Accuracy:",accuracy,"\n",
    "MSE:",MSE,"\n",
    "R-Sq. :",r2, "\n")
```

Now that we have a baseline for postseason prediction
