---
title: "College Football Regression"
author: "Blake Behrens"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
scores = read.csv('scores.csv')
scores$neutral = 0
scores$neutral[c(1,429,596,95,51,20,872,27)] = 1
postszn = read.csv("bowls25.csv")
```

# Linear Regression in College Football Prediction

The use of linear regression models in sports prediction is well documented and proved as a consistent and overall reliable model. The goal of this project is to analyze which predictors do best in forecasting the margin of scores between home team and away team. For this research, the regular season data will be built to train the model which will then be evaluated on playoff and bowl games (2025 season only).

## Basic Model

The first, most basic model to be fit is a multiple linear regression where observed score margin is explained by a difference in 2 team strengths. 

$$M_{ij} = S_i - S_j + \epsilon_{ij} $$
Where:

- $M_{ij}$ is the observed scoring margin for team *i* playing at home against team *j*

- $S_i,S_j$ are strengths of team *i* and *j* respectively

- $\epsilon_{ij}$ is the unobserved, mean-zero random error for the game between *i* and *j*

### Rankings: 

Below is this model's top 10 rankings, with the strength estimate representing the expected margin of victory against a true average team.


```{r Basic Model set-up, echo=FALSE}
basic.df = scores[scores$FBS == 1,c("home","away","margin")]
N = nrow(basic.df)
teams = unique(basic.df$home)
nteams = length(teams)
home = basic.df$home
away = basic.df$away

X = matrix(0, nrow = N, ncol = nteams)
i = 1
for(team in teams){
X[,i] = (home == team) -
  (away == team)
i = i + 1
}

y = basic.df$margin
basic.model = lm(y ~ X + 0)
b = coef(basic.model)
b = b - mean(b)
basic.ranks = data.frame("Team" = teams, "Strength" = b, row.names = 1:nteams)
head(arrange(basic.ranks,-Strength),10)
```


### Playoff Accuracy

```{r echo = F}
basic.evaluation = postszn %>% 
  mutate(margin = h.score - a.score) %>% 
  left_join(basic.ranks, by = join_by(home == Team)) %>% 
  dplyr::select(home,away,margin,homestrength = Strength) %>% 
  left_join(basic.ranks, by = join_by(away == Team)) %>% 
  dplyr::select(home,away,margin,homestrength, awaystrength = Strength) %>% 
  mutate(pred = homestrength - awaystrength) %>% 
  filter(!is.na(homestrength))

accuracy = mean(sign(basic.evaluation$margin) == sign(basic.evaluation$pred))
MSE = mean((basic.evaluation$margin - basic.evaluation$pred)^2)
r2 = cor(basic.evaluation$margin, basic.evaluation$pred)^2
n = nrow(basic.df)
p = length(basic.model$coefficients)
rss = sum(basic.model$residuals)
AIC <- n * log(rss / n) + 2 * p
BIC <- n * log(rss / n) + p * log(n)
cat("Basic Model: ", "\n",
    "Accuracy:",accuracy,"\n",
    "MSE:",MSE,"\n",
    "R-Sq. :",r2, "\n",
    "AIC:",AIC,"\n",
    "BIC:", BIC,"\n")
```

Now that there is a baseline for postseason prediction, I will construct some higher order models. I will incorporate various indicator variables referencing states about the game. This includes site, AP rank, and more.

## Model 2: Home Field Advantage

The first additional parameter I will investigate is adding an indicator for home field versus neutral site games. This will use the same model as before, 

$$M_{ij} = H + S_i - S_j + \epsilon_{ij} $$

but with the added $H$ term representing the effect of a game being played on a non-neutral site. 

```{r echo=F}
hfa.df = scores[scores$FBS == 1,c("home","away","margin","neutral")]
hfa.df$homegame = 1 - hfa.df$neutral
hfa.df$neutral = NULL
N = nrow(hfa.df)
teams = unique(hfa.df$home)
nteams = length(teams)
home = hfa.df$home
away = hfa.df$away

X = matrix(0, nrow = N, ncol = nteams+1) # +1 for adding homegame parameter
i = 2
for(team in teams){
X[,i] = (home == team) -
  (away == team)
i = i + 1
}
X[,1] = hfa.df$homegame

y = hfa.df$margin
hfa.model = lm(y ~ X + 0)
b = coef(hfa.model)[-1]
b = b - mean(b)
H = coef(hfa.model)[1]
hfa.ranks = data.frame("Team" = teams, "Strength" = b, row.names = 1:nteams)
head(arrange(hfa.ranks,-Strength),10)
```

As expected, since most games are not played on a neutral site, the estimated team strengths did not change much, and the top 10 remained the same. Hopefully, adding the term will help reduce our error in margin prediction, as we can expect it won't change the accuracy by much.

### Playoff Accuracy

```{r echo=F}
hfa.evaluation = postszn %>% 
  mutate(margin = h.score - a.score) %>% 
  left_join(hfa.ranks, by = join_by(home == Team)) %>% 
  dplyr::select(home,away,margin,neutral,homestrength = Strength) %>% 
  left_join(hfa.ranks, by = join_by(away == Team)) %>% 
  dplyr::select(home,away,margin,neutral,homestrength, awaystrength = Strength) %>% 
  mutate(homegame = 1- neutral,
         pred = homestrength - awaystrength + (H * homegame)) %>% 
  filter(!is.na(homestrength))

accuracy.h = mean(sign(hfa.evaluation$margin) == sign(hfa.evaluation$pred))
MSE.h = mean((hfa.evaluation$margin - hfa.evaluation$pred)^2)
r2.h = cor(hfa.evaluation$margin, hfa.evaluation$pred)^2
n.h = nrow(hfa.df)
p.h = length(hfa.model$coefficients)
rss.h = sum(hfa.model$residuals)
AIC.h <- n.h * log(rss.h / n.h) + 2 * p.h
BIC.h <- n.h * log(rss.h / n.h) + p.h * log(n.h)
data.frame("Basic_Model" = c(accuracy,MSE,r2,AIC,BIC),
           "Home_Field_Model" = c(accuracy.h,MSE.h,r2.h,AIC.h,BIC.h),
           row.names = c("Accuracy","MSE","R-Squared","AIC","BIC"))
```

To my surprise, adding the term for a "home field advantage" reduces model accuracy and increases the mean squared error. My assumption is that since the model is trained on 99.1% of non-neutral games, the estimate for the home field advantage parameter is biased. Therefore, it does not perform well on post-season games where only 10.8% are played at home. I will also note that of the 5 post season games played with a home field advantage, this model correctly picked the winner 4 times, with mean squared error of 94.5. In summary, a model like this would perhaps perform better in NFL post-season prediction where only 1 game (Super Bowl) can be played at a neutral site.

## Model 3: AP Ranks as factors

Now, I will investigate how AP rankings can assist the linear model. I will no longer be including the home field parameter model. Further, since AP ranks teams 1-25 and therefore most teams don't have an assigned rank, I will instead use indicators that say whether or not AP ranked that team:

$$M_{ij} = S_i - S_j + [I(\text{AP}_i) * A] - [I(\text{AP}_j) * A] + \epsilon_{ij}$$

or:

$$M_{ij} = S_i - S_j + A * (I(\text{AP}_i) - I(\text{AP}_j)) + \epsilon_{ij}$$

Where:

- $I(\text{AP}_i)$ and $I(\text{AP}_j)$ are indicator functions of whether team *i* and *j* are ranked, respectively

- $A$ is the effect on the margin from a team being ranked on the AP poll

```{r echo=F}
ap.df = scores[,c("home","away",'margin','home_AP','away_AP')]
ap.df$home_AP = as.numeric(!is.na(ap.df$home_AP))
ap.df$away_AP = as.numeric(!is.na(ap.df$away_AP))
ap.df$AP.diff = ap.df$home_AP - ap.df$away_AP

N = nrow(ap.df)
teams = unique(ap.df$home)
nteams = length(teams)
home = ap.df$home
away = ap.df$away

X = matrix(0, nrow = N, ncol = nteams+1)
i = 2
for(team in teams){
  X[,i] = (home == team) - (away == team)
  i = i + 1
}

X[,1] = ap.df$AP.diff
y = ap.df$margin
ap.model = lm(y ~ X + 0)

b = coef(ap.model)[-1]
AP = coef(ap.model)[1]

b = b - mean(b)
ap.ranks = data.frame("Team" = teams, "Strength" = b, row.names = 1:nteams)
head(arrange(ap.ranks,-Strength),10)
```

Here we notice that the top 10 is different (excluding the top 3 teams, Indiana, Ohio State, and Texas Tech). It appears AP rank had more impact on the parameter estimation than including homefield advantage did.



```{r echo=F}
ap.postszn = data.frame(
    Team = c(
    'IU','UGA','OSU','TTU','ORE',
    'MISS','TA&M','OU','ND','MIA',
    'ALA','BYU','VAN','TEX','UTAH',
    'USC','TULN','MICH','JMU','UVA',
    'ARIZ','NAVY','UNT','GT','MIZ'
  ),
  Rank = 1:25
)

ap.evaluation = postszn %>% 
  mutate(margin = h.score - a.score) %>% 
  left_join(ap.ranks, by = join_by(home == Team)) %>% 
  dplyr::select(home,away,margin,neutral,homestrength = Strength) %>% 
  left_join(hfa.ranks, by = join_by(away == Team)) %>% 
  dplyr::select(home,away,margin,neutral,homestrength, awaystrength = Strength) %>% 
  mutate(homegame = 1- neutral,
         pred = homestrength - awaystrength + (H * homegame)) %>% 
  filter(!is.na(homestrength)) %>% 
  left_join(ap.postszn, by = join_by(home == Team)) %>% 
  dplyr::select(home,away,margin,neutral,homestrength, awaystrength, home.ap = Rank) %>% 
  left_join(ap.postszn, by = join_by(away == Team)) %>% 
  dplyr::select(home,away,margin,neutral,homestrength, awaystrength, home.ap, away.ap = Rank) %>% 
  mutate(
    home.ap = as.numeric(!is.na(home.ap)),
    away.ap = as.numeric(!is.na(away.ap)),
    AP.diff = home.ap - away.ap,
    pred = homestrength - awaystrength + (AP * AP.diff)
  )

accuracy.a = mean(sign(ap.evaluation$margin) == sign(ap.evaluation$pred))
MSE.a = mean((ap.evaluation$margin - ap.evaluation$pred)^2)
r2.a = cor(ap.evaluation$margin, ap.evaluation$pred)^2
n.a = nrow(ap.df)
p.a = length(ap.model$coefficients)
rss.a = sum(ap.model$residuals)
AIC.a <- n.a * log(rss.a / n.a) + 2 * p.a
BIC.a <- n.a * log(rss.a / n.a) + p.a * log(n.a)
data.frame("Basic_Model" = c(accuracy,MSE,r2,AIC,BIC),
           "Home_Field_Model" = c(accuracy.a,MSE.a,r2.a,AIC.a,BIC.a),
           row.names = c("Accuracy","MSE","R-Squared","AIC","BIC"))
```